Appendix 1: sdmTMB statistical model description

### 2022-03-16 doc/appendix-model.Rmd
Edits to model appendix

This appendix uses the following notation conventions, generally following the guidance in @edwards2019:

@@

$$
\begin{aligned}
\mathbb{E}[y_{\s,t}] &= \mu_{\s,t},\\
\mu_{\bm{s},t} &=
f^{-1} \left( \bm{X}_{1,\bm{s},t} \bm{\beta} +
O_{\s,t} +
\alpha_g +
\bm{X}_{2,\bm{s},t} \gamma_t +
\bm{X}_{3,t} \zeta_{\bm{s}} +
\omega_{\s} +
\epsilon_{\s,t} \right),
\end{aligned}
$$

@@

A single sdmTMB model will rarely, if ever, contain all of the above components. Next, we will split the model apart to describe the various parts in more detail using '$\ldots$' to represent the other optional components.

@@

Smoothers in sdmTMB are implemented with the same formula syntax familiar to mgcv users fitting GAMs (generalized additive models).
Smooths are implemented in the formula using `+ s(x)`, which implements a smooth from `mgcv::s()`.
Within these smooths, the same syntax commonly used in `mgcv::s()` can be applied, e.g. 2-dimensional smooths may be constructed with `+ s(x, y)`; smooths can be specific to various factor levels, `+ s(x, by = group)`; smooths can vary according to a continuous variable, `+ s(x, by = x2)`; the basis function dimensions may be specified, e.g. `+ s(x, k = 4)`; and various types of splines may be constructed such as cyclic splines to model seasonality, e.g. `+ s(month, bs = "cc", k = 12)`.

@@

These models can be fit by including `+ breakpt(x)` in the model formula, where
`x` is a covariate.

@@

These models can be fit by including `+ logistic(x)` in the model formula, where `x` is a covariate.

@@

Spatial random fields, $\omega_{\s}$, are included if `spatial = 'on'` and omitted if `spatial = 'off'`.

$$
\begin{aligned}
\mu_{\bm{s},t} &= f^{-1} \left( \ldots + \omega_{\s} + \ldots \right),\\
\boldsymbol{\omega} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),\\
\end{aligned}
$$
The marginal standard deviation of $\boldsymbol{\omega}$ is indicated by `Spatial SD` in the model output or as `sigma_O` in the output of `sdmTMB::tidy(fit, "ran_pars")`.

@@

Spatiotemporal random fields are represented by $\boldsymbol{\epsilon}_t$ within sdmTMB.
This has been chosen to match the representation in VAST [@thorson2019].
The marginal standard deviation of $\boldsymbol{\epsilon}_t$ is indicated by `Spatiotemporal SD` in the model output or as `sigma_E` in the output of `sdmTMB::tidy(fit, "ran_pars")`.

@@

where $\epsilon_{\s,t}$ represent random field deviations at point $\s$ and time $t$.
The random fields are assumed independent across time slices.

@@

First-order auto regressive, AR(1), spatiotemporal random fields (`spatiotemporal = 'ar1'`) add a parameter defining the correlation between random field deviations from one time slice to the next.
They are defined as

@@

Random walk spatiotemporal random fields (`spatiotemporal = 'rw'`) represent a model where the difference in spatiotemporal deviations from one time slice to the next are IID. There defined as

@@

Spatially varying coefficient (SVC) models are defined as

@@

@@

TMB allows for anisotropy, where spatial covariance may be asymmetric with respect to latitude and longitude ([full details](https://kaskr.github.io/adcomp/namespaceR__inla.html)). Anisotropy can be turned on and off with the logical `anisotropy` argument to `sdmTMB()`. There are a number of ways to implement anisotropic covariance [@fuglstad2015a], and we adopt a 2-parameter rotation matrix $\textbf{H}$. The elements of $\textbf{H}$ are defined by the parameter vector $\boldsymbol{x}$ so that $H_{1,1} = x_{1}$, $H_{1,2} = H_{2,1} = x_{2}$ and $H_{2,2} = (1 + x_{2}^2) / x_{1}$.

@@

In some cases the spatial domain of interest may be complex and bounded by some barrier such as by land or water (e.g., coastlines, islands, lakes).
SPDE models allow for physical barriers to be incorporated into the modelling [@bakka2019].
With  `sdmTMB()` models, the mesh construction occurs in two steps: the user (1) constructs a mesh with a call to `sdmTMB::make_mesh()`, and (2) passes the mesh to `sdmTMB::add_barrier_mesh()`.
The barriers must be constructed as `sf` objects [@pebesma2018] with polygons defining the barriers.
See `?sdmTMB::add_barrier_mesh` for examples.

>>>>

This appendix uses the following notation conventions, which generally follows the guidance in @edwards2019:

@@

$$
\begin{aligned}
\mathbb{E}[y_{\s,t}] &= \mu_{\s,t},\\
\mu_{\bm{s},t} &=
f^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
O_{\s,t} +
\alpha_g +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \gamma_t +
\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
\omega_{\s} +
\epsilon_{\s,t} \right),
\end{aligned}
$$

@@

A single sdmTMB model will rarely, if ever, contain all of the above components. Next, we will split the model to describe the various parts in more detail using '$\ldots$' to represent the other optional components.

@@

Smoothers in sdmTMB are implemented with the same formula syntax familiar to mgcv [@wood2017a] users fitting GAMs (generalized additive models).
Smooths are implemented in the formula using `+ s(x)`, which implements a smooth from `mgcv::s()`.
Within these smooths, the same syntax commonly used in `mgcv::s()` can be applied, e.g. 2-dimensional smooths may be constructed with `+ s(x, y)`; smooths can be specific to various factor levels, `+ s(x, by = group)`; smooths can vary according to a continuous variable, `+ s(x, by = x2)`; the basis function dimensions may be specified, e.g. `+ s(x, k = 4)` (see `?mgcv::choose.k`); and various types of splines may be constructed such as cyclic splines to model seasonality, e.g. `+ s(month, bs = "cc", k = 12)`.

@@

These models can be fit by including `+ breakpt(x)` in the model formula, where
`x` is a covariate.
The formula can contain a single break-point covariate.

@@

These models can be fit by including `+ logistic(x)` in the model formula, where `x` is a covariate.
The formula can contain a single logistic covariate.

@@

Spatial random fields, $\omega_{\s}$, are included if `spatial = 'on'` (or `TRUE`) and omitted if `spatial = 'off'` (or `FALSE`).

$$
\begin{aligned}
\mu_{\bm{s},t} &= f^{-1} \left( \ldots + \omega_{\s} + \ldots \right),\\
\boldsymbol{\omega} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),\\
\end{aligned}
$$
The marginal standard deviation of $\boldsymbol{\omega}$ is indicated by `Spatial SD` in the printed model output or as `sigma_O` in the output of `sdmTMB::tidy(fit, "ran_pars")`. The 'O' is for 'omega' ($\omega$).

@@

Spatiotemporal random fields are represented by $\boldsymbol{\epsilon}_t$ within sdmTMB.
This has been chosen to match the representation in VAST [@thorson2019].
The marginal standard deviation of $\boldsymbol{\epsilon}_t$ is indicated by `Spatiotemporal SD` in the printed model output or as `sigma_E` in the output of `sdmTMB::tidy(fit, "ran_pars")`.
The 'E' is for 'epsilon' ($\epsilon$).

@@

where $\epsilon_{\s,t}$ represent random field deviations at point $\s$ and time $t$.
The random fields are assumed independent across time steps.

@@

First-order auto regressive, AR(1), spatiotemporal random fields (`spatiotemporal = 'ar1'`) add a parameter defining the correlation between random field deviations from one time step to the next.
They are defined as

@@

Random walk spatiotemporal random fields (`spatiotemporal = 'rw'`) represent a model where the difference in spatiotemporal deviations from one time step to the next are IID. They are defined as

@@

Spatially varying coefficient models are defined as

@@

Multilevel/hierchical intercepts are defined as

@@

TMB allows for anisotropy, where spatial covariance may be asymmetric with respect to latitude and longitude ([full details](https://kaskr.github.io/adcomp/namespaceR__inla.html)). Anisotropy can be turned on or off with the logical `anisotropy` argument to `sdmTMB()`. There are a number of ways to implement anisotropic covariance [@fuglstad2015a], and we adopt a 2-parameter rotation matrix $\textbf{H}$. The elements of $\textbf{H}$ are defined by the parameter vector $\boldsymbol{x}$ so that $H_{1,1} = x_{1}$, $H_{1,2} = H_{2,1} = x_{2}$ and $H_{2,2} = (1 + x_{2}^2) / x_{1}$.

@@

In some cases the spatial domain of interest may be complex and bounded by some barrier such as by land or water (e.g., coastlines, islands, lakes).
SPDE models allow for physical barriers to be incorporated into the modelling [@bakka2019].
With  `sdmTMB()` models, the mesh construction occurs in two steps: the user (1) constructs a mesh with a call to `sdmTMB::make_mesh()`, and (2) passes the mesh to `sdmTMB::add_barrier_mesh()`.
The barriers must be constructed as `sf` objects [@pebesma2018] with polygons defining the barriers.
See `?sdmTMB::add_barrier_mesh` for an example.
